\stktokens{}: Enforcing Well-Bracketed Control Flow and Stack Encapsulation Using Linear Capabilities

### 2019-05-21 popl-2019-journal-v/proving-full-abstraction.tex
[POPLjv] Update todos

% Closure seals
Unlike return seals, both trusted and untrusted components can have closure seals.
For untrusted components (components with their code address space disjoint from the trusted address space), we allow everything in the untrusted value relation to be sealed.
Intuitively, untrusted components are assumed to have access to words from the untrusted value relation, but we cannot know how the words are used, so we need to assume that they may be sealed.
% TODO trusted components
Trusted components only use closure seals for sealed capability pairs that represent actual closures.
The code capability for a closure must point to the code memory because it is the only part of memory that is executable.
It is not safe for untrusted components to have a capability for a trusted components code (it could be used to read return capabilities or start execution in the middle of a call), so capabilities for the code memory of a trusted component is in the trusted value relation.
While it is not safe to give a bare capability for a trusted components code memory, it can be perfectly safe to give a sealed capability for a trusted components code.
For this reason, the seal invariant allows executable capabilities from the trusted value relation to be sealed with a closure seal.
When it comes to the data capability of a closure, we just require that it comes from the untrusted value relation because the trusted value relation contains nothing that makes sense to seal as the data capability (we return to the specific contents of the two value relations in a later paragraph).

@@

@@

\paragraph{The trusted value relation}
\label{par:trusted-val-rel}
\[
  \lrvtrusted[\square,\gc](W) =
  \begin{array}[t]{l}
    \lrv(W)\cup \\
%
    \left\{ \npair{\left(\arraycolsep=0pt\array{l} \src{\seal{\sigma_\baddr,\sigma_\eaddr,\sigma}},\\ \seal{\sigma_\baddr,\sigma_\eaddr,\sigma} \endarray \right)} 
    \; \middle| \;
    \begin{array}{l}
      \gc = (\ta,\stkb,\gsigrets,\gsigcloss)  \wedge \\
      \exists r \in \dom(\pwheap) \ldotp \\
      \quad \pwheap(r) \nequal \codereg{\sigrets,\sigcloss,\code,(\ta,\stkb,\gsigrets,\gsigcloss)} \wedge\\
      \quad \dom(\code) \subseteq \ta \wedge [\sigma_\baddr,\sigma_\eaddr] \subseteq (\sigrets\cup\sigcloss) \wedge\\
      \quad \sigrets \subseteq \gsigrets \wedge \sigcloss \subseteq \gsigcloss
    \end{array}
    \right\} \cup \\
    \left\{ \npair{\left(\arraycolsep=0pt\array{l} \src{((\perm,\normal),\baddr,\eaddr,\aaddr)},\\ ((\perm,\normal),\baddr,\eaddr,\aaddr) \endarray \right)} \;\middle|\; 
    \begin{array}{l}
      \perm \sqsubseteq \rx \wedge \\
      \gc = (\ta,\stkb,\gsigrets,\gsigcloss)  \wedge \\
      {} [\baddr,\eaddr] \subseteq \ta \wedge\\
      \npair{[\baddr,\eaddr]} \in \xReadCond[\square,\gc]{W} 
    \end{array}
    \right\}
  \end{array}
\]

@@

\subsubsection{Expression relations}
% Two expression relations: one for sealed code-data pairs being jumped to and one for capabilities being jumped to in the regular way.
% The argument for having one relation relate pairs of pairs of capabilities and the other relate pairs of capabilities is that that is how xjump and regular jumps work: xjump takes pairs while regular jumps take single capabilities.
\begin{align*}
  \lrexj(W) &= \left\{ \npair{\stpair[.]{v_{c,S},v_{d,S}}{v_{c,T},v_{d,T}}} \middle| 
    \begin{array}{l}
      \forall n' \leq n, \src{\reg_S}, \reg_T, \src{\ms_S}, \ms_T, \src{\ms_\stk}, \src{\stk} \ldotp\\
      \quad \forall W_\lrrs , W_\lrm \ldotp \\
      \qquad\npair[n']{\stpair{\reg}{\reg}} \in \lrr(\{\rdata\}) (W_\lrrs ) \wedge\\
      \qquad\memSat[n']{\stpair[.]{\ms_S,\stk,\ms_\stk}{\ms_T}}{W_\lrm} \wedge \\
      \qquad\Phi_S = \src{(\ms_S,\reg_S,\stk, \ms_\stk)}\wedge\\
      \qquad\Phi_T = (\ms_T,\reg_T) \wedge\\
      \qquad W \oplus W_\lrrs \oplus W_\lrm \text{ is defined }\\
      \qquad\Rightarrow \exists \Phi_S',\Phi_T'\ldotp\\
      \quad\qquad \Phi_S' = \xjumpResult{v_{c,S}}{v_{d,S}}{\Phi_S} \tand\\
      \quad\qquad\Phi_T' = \xjumpResult{v_{c,T}}{v_{d,T}}{\Phi_T}\tand\\
      \quad\qquad\npair[n']{\left(\Phi_S', \Phi_T' \right)}\in \lro
    \end{array}
    \right\}\\
  \lre(W) &= \left\{ \npair{\stpair[.]{v_{c,S}}{v_{c,T}}} \middle| 
    \begin{array}{l}
      \forall n' \leq n, \src{\reg_S}, \reg_T, \src{\ms_S}, \ms_T, \src{\ms_\stk}, \src{\stk} \ldotp\\
      \quad \forall W_\lrrs , W_\lrm \ldotp \\
      \qquad\npair[n']{\stpair{\reg}{\reg}} \in \lrr(W_\lrrs ) \wedge\\
      \qquad\memSat[n']{\stpair[.]{\ms_S,\stk,\ms_\stk}{\ms_T}}{W_\lrm} \\
      \qquad\Phi_S = \src{(\ms_S,\reg_S,\stk, \ms_\stk)}\\
      \qquad \Phi_S' = \Phi_S \updReg{\pcreg}{v_{c,S}}\\
      \qquad\Phi_T = (\ms_T,\reg_T)\\
      \qquad\Phi_T' = \Phi_T\updReg{\pcreg}{v_{c,T}}\\
      \qquad W \oplus W_\lrrs \oplus W_\lrm\\
      \qquad\Rightarrow\npair[n']{\left(\Phi_S', \Phi_T' \right)}\in \lro
    \end{array}
    \right\}
\end{align*}
\end{jversion}
\subsection{Fundamental Theorem}
An important lemma in our proof of full abstraction of the embedding of \srccm{} into \trgcm{}, is the fundamental theorem of logical relations (FTLR).
The name indicates that it is an instance of a general pattern in logical relations proofs, but is otherwise unimportant.
\begin{lemma}[FTLR (roughly)]
  \label{thm:ftlr}
  If $[\baddr,\eaddr] \subseteq \dom(\mscode)$ and $\pwheap(r) = \codereg{\sigrets,\sigcloss,\mscode}$, and either $[\baddr,\eaddr] \subseteq \ta$ and $\mscode$ behaves reasonably (see Section~\ref{sec:well-form-reas}) or
$[\baddr,\eaddr] \mathrel{\#} \ta$,
then 
  \[
    \npair{\left(\src{((\rx,\normal),\baddr,\eaddr,\aaddr)}, ((\rx,\normal),\baddr,\eaddr,\aaddr)\right)} \in \lre(W) \qedhere
  \]
\end{lemma}

@@

>>>>

% Closure seals
Unlike return seals, both trusted and untrusted components can have closure seals.
For untrusted components (components with their code address space disjoint from the trusted address space), we allow everything in the untrusted value relation to be sealed.
Intuitively, untrusted components are assumed to have access to words from the untrusted value relation, but we cannot know how the words are used, so we need to assume that they may be sealed.
Trusted components only use closure seals for sealed capability pairs that represent actual closures.
The code capability for a closure must point to the code memory because it is the only part of memory that is executable.
It is not safe for untrusted components to have a capability for a trusted components code (it could be used to read return capabilities or start execution in the middle of a call), so capabilities for the code memory of a trusted component is in the trusted value relation.
While it is not safe to give a bare capability for a trusted components code memory, it can be perfectly safe to give a sealed capability for a trusted components code.
For this reason, the seal invariant allows executable capabilities from the trusted value relation to be sealed with a closure seal.
When it comes to the data capability of a closure, we just require that it comes from the untrusted value relation because the trusted value relation contains nothing that makes sense to seal as the data capability (we return to the specific contents of the two value relations in a later paragraph).

@@

% TODO write section

@@

\paragraph{The trusted value relation}
\label{par:trusted-val-rel}
% TODO write section
\[
  \lrvtrusted[\square,\gc](W) =
  \begin{array}[t]{l}
    \lrv(W)\cup \\
%
    \left\{ \npair{\left(\arraycolsep=0pt\array{l} \src{\seal{\sigma_\baddr,\sigma_\eaddr,\sigma}},\\ \seal{\sigma_\baddr,\sigma_\eaddr,\sigma} \endarray \right)} 
    \; \middle| \;
    \begin{array}{l}
      \gc = (\ta,\stkb,\gsigrets,\gsigcloss)  \wedge \\
      \exists r \in \dom(\pwheap) \ldotp \\
      \quad \pwheap(r) \nequal \codereg{\sigrets,\sigcloss,\code,(\ta,\stkb,\gsigrets,\gsigcloss)} \wedge\\
      \quad \dom(\code) \subseteq \ta \wedge [\sigma_\baddr,\sigma_\eaddr] \subseteq (\sigrets\cup\sigcloss) \wedge\\
      \quad \sigrets \subseteq \gsigrets \wedge \sigcloss \subseteq \gsigcloss
    \end{array}
    \right\} \cup \\
    \left\{ \npair{\left(\arraycolsep=0pt\array{l} \src{((\perm,\normal),\baddr,\eaddr,\aaddr)},\\ ((\perm,\normal),\baddr,\eaddr,\aaddr) \endarray \right)} \;\middle|\; 
    \begin{array}{l}
      \perm \sqsubseteq \rx \wedge \\
      \gc = (\ta,\stkb,\gsigrets,\gsigcloss)  \wedge \\
      {} [\baddr,\eaddr] \subseteq \ta \wedge\\
      \npair{[\baddr,\eaddr]} \in \xReadCond[\square,\gc]{W} 
    \end{array}
    \right\}
  \end{array}
\]

@@

\subsubsection{Expression relations}
% TODO write section
% Two expression relations: one for sealed code-data pairs being jumped to and one for capabilities being jumped to in the regular way.
% The argument for having one relation relate pairs of pairs of capabilities and the other relate pairs of capabilities is that that is how xjump and regular jumps work: xjump takes pairs while regular jumps take single capabilities.
\begin{align*}
  \lrexj(W) &= \left\{ \npair{\stpair[.]{v_{c,S},v_{d,S}}{v_{c,T},v_{d,T}}} \middle| 
    \begin{array}{l}
      \forall n' \leq n, \src{\reg_S}, \reg_T, \src{\ms_S}, \ms_T, \src{\ms_\stk}, \src{\stk} \ldotp\\
      \quad \forall W_\lrrs , W_\lrm \ldotp \\
      \qquad\npair[n']{\stpair{\reg}{\reg}} \in \lrr(\{\rdata\}) (W_\lrrs ) \wedge\\
      \qquad\memSat[n']{\stpair[.]{\ms_S,\stk,\ms_\stk}{\ms_T}}{W_\lrm} \wedge \\
      \qquad\Phi_S = \src{(\ms_S,\reg_S,\stk, \ms_\stk)}\wedge\\
      \qquad\Phi_T = (\ms_T,\reg_T) \wedge\\
      \qquad W \oplus W_\lrrs \oplus W_\lrm \text{ is defined }\\
      \qquad\Rightarrow \exists \Phi_S',\Phi_T'\ldotp\\
      \quad\qquad \Phi_S' = \xjumpResult{v_{c,S}}{v_{d,S}}{\Phi_S} \tand\\
      \quad\qquad\Phi_T' = \xjumpResult{v_{c,T}}{v_{d,T}}{\Phi_T}\tand\\
      \quad\qquad\npair[n']{\left(\Phi_S', \Phi_T' \right)}\in \lro
    \end{array}
    \right\}\\
  \lre(W) &= \left\{ \npair{\stpair[.]{v_{c,S}}{v_{c,T}}} \middle| 
    \begin{array}{l}
      \forall n' \leq n, \src{\reg_S}, \reg_T, \src{\ms_S}, \ms_T, \src{\ms_\stk}, \src{\stk} \ldotp\\
      \quad \forall W_\lrrs , W_\lrm \ldotp \\
      \qquad\npair[n']{\stpair{\reg}{\reg}} \in \lrr(W_\lrrs ) \wedge\\
      \qquad\memSat[n']{\stpair[.]{\ms_S,\stk,\ms_\stk}{\ms_T}}{W_\lrm} \\
      \qquad\Phi_S = \src{(\ms_S,\reg_S,\stk, \ms_\stk)}\\
      \qquad \Phi_S' = \Phi_S \updReg{\pcreg}{v_{c,S}}\\
      \qquad\Phi_T = (\ms_T,\reg_T)\\
      \qquad\Phi_T' = \Phi_T\updReg{\pcreg}{v_{c,T}}\\
      \qquad W \oplus W_\lrrs \oplus W_\lrm\\
      \qquad\Rightarrow\npair[n']{\left(\Phi_S', \Phi_T' \right)}\in \lro
    \end{array}
    \right\}
\end{align*}
\end{jversion}
\subsection{Fundamental Theorem}
% TODO rewrite section
% Include the actual FTLR
An important lemma in our proof of full abstraction of the embedding of \srccm{} into \trgcm{}, is the fundamental theorem of logical relations (FTLR).
The name indicates that it is an instance of a general pattern in logical relations proofs, but is otherwise unimportant.
\begin{lemma}[FTLR (roughly)]
  \label{thm:ftlr}
  If $[\baddr,\eaddr] \subseteq \dom(\mscode)$ and $\pwheap(r) = \codereg{\sigrets,\sigcloss,\mscode}$, and either $[\baddr,\eaddr] \subseteq \ta$ and $\mscode$ behaves reasonably (see Section~\ref{sec:well-form-reas}) or
$[\baddr,\eaddr] \mathrel{\#} \ta$,
then 
  \[
    \npair{\left(\src{((\rx,\normal),\baddr,\eaddr,\aaddr)}, ((\rx,\normal),\baddr,\eaddr,\aaddr)\right)} \in \lre(W) \qedhere
  \]
\end{lemma}

@@

\subsection{Related Components}
% TODO write section
\begin{align*}
  \lrcomp(W) &=
  \left\{\begin{aligned}
     &\npair{\var{comp},\var{comp}} \;\mid \;\\
      &\qquad\var{comp} = (\mscode,\msdata,\overline{a_{\mathrm{import}} \mapsfrom s_{\mathrm{import}}},\overline{s_{\mathrm{export}} \mapsto w_{\mathrm{export}}},\sigrets,\sigcloss) \tand \\
      &\qquad\text{For all } W' \future W \ldotp \\
      &\qquad\quad\text{If } \overline{\npair[n']{(w_{\mathrm{import}},w_{\mathrm{import}})}} \in \lrv(\purePart{W'}) \text{ for all $n' < n$}\\
      &\qquad\quad\text{and } \msdata' = \msdata{}[\overline{a_{\mathrm{import}} \mapsto w_{\mathrm{import}}}] \\
      &\qquad\quad\text{then } \npair{(\sigrets\uplus\sigcloss,\mscode\uplus \msdata', \mscode\uplus\msdata')} \in \lrheap(\pwheap)(W') \tand\\
      &\qquad\quad\overline{\npair{(w_{\mathrm{export}},w_{\mathrm{export}})}} \in \lrv(\purePart{W'})
  \end{aligned}
    \right\}\\
  &\cup \left\{
    \begin{multlined}
     \npair{(\var{comp}_0,c_{\mathrm{main},c}, c_{\mathrm{main},d}),(\var{comp}_0,c_{\mathrm{main},c}, c_{\mathrm{main},d})} \;\mid \;\\
     \npair{(\var{comp}_0,\var{comp}_0)} \in \lrcomp(W) \tand\\
     \{(\_ \mapsto c_{\mathrm{main},c}),(\_ \mapsto c_{\mathrm{main},d})\} \subseteq \overline{w_{\mathrm{export}}}
       \end{multlined}
    \right\} 
\end{align*}
% TODO include central lemmas?

\subsection{Related Execution Configuration}
% TODO write section
\[
  \lrec(W) = \left\{ \npair{\left(
        \arraycolsep=0pt\array{l}
        (\ms_S,\reg_S,\stk,\ms_\stk),\\
        (\ms_T,\reg_T)\endarray\right)} \middle|
    \begin{array}{l}
      \gc = (\ta,\stkb) \tand\\
      \exists W_M,W_R,W_\pcreg \ldotp W = W_M \oplus W_R \oplus W_\pcreg \tand\\
      \npair{( (\reg_S(\pcreg),\reg_S(\rdata)), (\reg_T(\pcreg),\reg_T(\rdata)) )} \in \lrexj(W_\pcreg) \tand \\
      \reg_S(\pcreg) \neq \retptrc(\_) \wedge 
      \reg_S(\rdata) \neq \retptrd(\_) \wedge \\
      \nonExec{\reg_S(\rdata)} \wedge
      \nonExec{\reg_T(\rdata)}\\
      \memSat{\ms_S,\ms_\stk,\stk,\ms_T}{W_M} \tand\\
      \npair{\stpair{\reg}{\reg}} \in \lrr(\{\rdata\})(W_R)
    \end{array}
            \right\}
\]

% TODO Include central lemmas?
